Missing the Point\(\colon\) Non-Convergence in Iterative Imputation Procedures

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.Rmd
Include the acf into simulation again to show differences

Missing data problems are ubiquitous and pervasive. **or** Everybody who works with person data is likely to run into a missing data problem.  
<!-- At some point, any social scientist  -->
<!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  -->
Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18]. A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76]. The multiple imputation method can be roughly divided into 3 steps: First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times by some algorithmic procedure. Next, the analysis of scientific interest is performed on each of the completed datasets. And finally, the results of these analyses are pooled into a single aggregated statistic.

@@

Algorithms are often used to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data?

@@

There is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully. The current practice is visually inspecting imputation chains for signs of non-convergence, but this may be challenging to the untrained eye [@buur18, $\S$ 6.5.2]. Moreover, visually inspecting imputation chains may only diagnose severely pathological cases of non-convergence. Therefore, diagnostic evaluation of convergence would be preferred.

Iterative imputation algorithms such as mice, are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another.  Therefore, generated values will vary even after convergence. The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose severe non-convergence.
<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

Several MCMC convergence diagnostics exist, but it is not known whetherthese are appropriate for MICE. The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. The open questions are, e.g.: How can non-convergence be diagnosed? How many iterations are sufficient/needed? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

>>>>

<!-- (**or** Everybody who works with person data is likely to run into a missing data problem.) -->
Missing data problems are ubiquitous and pervasive in the social and behavioral sciences. If a dataset contains just one missing value for a variable of interest, statistical inferences are undefined and will not produce any result. Incomplete observations are therefore ignored by default in many statistical packages (i.e., list-wise deletion is employed). Unfortunaltely, this *ad hoc* solution may yield wildly invalid results [@buur18]. An alternative is to 'impute' (i.e., fill in) every missing value in an incomplete dataset. With imputation techniques, one or several completed datasets are created, on which statistical inferences can be performed. The case with several completed datasets is known as 'multiple imputation' [MI; @rubin76], and requires an additional step to pool the results accross imputations. By applying Rubin's rules, the pooled result will reflect the uncertainty in the data due to missingness. Under many circumstances, this yields an unbiased and confidence valid estimate of the inference on the true--but missing--data [@buur18]. A popular method to obtain imputations is to use the 'Multiple Imputation by Chained Equations' algorithm, shorthand 'MICE'. MICE is an iterative algorithmic procedure to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data? These are all open questions, because the convergence properties of iterative imputation algortihms have not been systematically studied [@taka17; @buur18]. In this paper, we investigate the convergence properties of iterative imputation procedures by means of model based simulation in `R` [@R].

<!-- It has been shown that this pooled result MI will yield unbiased, confidence valid results under many missing data circumstances [@buur18]. To implement MI,  -->

<!-- Rubin's rules [-@rubin76] may be used to reflect the uncertainty in the inferences due to missingness by pooling the results across 'imputations'. A popular method to obtain imputations   completions, the analysis of scientific interest can be performed. It has been , or to complete these observations by imputing the missing values. Imputation is the method of filling in  -->

<!-- by which  That is, unless ad hoc solutions are employed (e.g., list-wise deletion), which can yield invalid results.  contains missing values, statistical analyses cannot be performed without  only be performed a dataset contains missing values, statistical analyses can statistical inferences cannot be performed   -->
<!-- <!-- At some point, any social scientist  --> -->
<!-- <!-- conducting statistical analyses  --> <!-- will run into a missing data problem [@alli02].  --> -->
<!-- Missingness is problematic because statistical inference cannot be performed on the whole data without employing \emph{ad hoc} solutions (e.g., list-wise deletion), which may yield wildly invalid results [@buur18]. A popular method to obtain statistically valid results despite of missing data is 'multiple imputation' [MI; @rubin76]. The multiple imputation method can be roughly divided into 3 steps: First, each missing value in an incomplete dataset is 'imputed' (i.e., filled in) several times by some algorithmic procedure. Next, the analysis of scientific interest is performed on each of the completed datasets. And finally, the results of these analyses are pooled into a single aggregated statistic.  -->

@@

<!-- Algorithms are often used to draw imputations from the posterior predictive distribution of the missing values. This introduces a potential threat to the validity of the imputations: what if the algorithm has not converged? Are the implications then to be trusted? And can we rely on the inference obtained on the completed data?  -->

@@

**(There is no scientific consensus on the convergence properties of MI algorithms [@taka17]. Some default MICE techniques (e.g., 'predictive mean modeling') might not yield converged states at all [@murr18]. Therefore, algorithmic convergence should be monitored carefully.)**

The recommended practice to monitor convergence is to visually inspect imputation chains for signs of non-convergence. This practice is insuffcient on two counts: 1) it may be challenging to the untrained eye [@buur18, $\S$ 6.5.2], and 2) only severely pathological cases of non-convergence may be diagnosed. Therefore, a quantitative, diagnostic evaluation of convergence would be preferred.

Iterative imputation algorithms such as MICE are MCMC methods, and convergence of MCMC algorithms is not from a scalar to a point but from one distribution to another. Therefore, generated values will vary even after convergence. The aim of convergence diagnostics for MCMC methods is thus not to establish the point at which convergence is reached, but to diagnose signs of  non-convergence.
<!-- Therefore we can only diagnose non-convergence, not convergence [@hoff09]. "a weak diagnostics is better than no diagnostic at all" [@cowl96]. -->

Several MCMC convergence diagnostics exist, but it is not known whether these are appropriate for MICE. The application of convergence diagnostics to MI methods has not been systematically studied [@buur18]. The open questions are, e.g.: How can non-convergence be diagnosed? How many iterations are sufficient/needed? What are the effects of non-convergence on estimates, predictions and inference? What is an appropriate non-convergence diagnostic? Are the default number of iterations sufficient? The defaults of common MI software: SPSS = 10 [(link)](https://www.ibm.com/support/knowledgecenter/SSLVMB_24.0.0/spss/mva/syn_multiple_imputation_impute.html), Mplus = 100?? [(link)](https://pdfs.semanticscholar.org/e20e/29e008592cbfbaa567931f74cdfdb5451405.pdf?_ga=2.55354671.54033656.1584698748-527613517.1584698748), Stata = 10 [(link)](https://www.stata.com/manuals13/mi.pdf, p. 139), Amelia = NA, because resampling, not convergence [(link)](https://cran.r-project.org/web/packages/Amelia/Amelia.pdf), en MI = 30 [(link)](https://cran.r-project.org/web/packages/mi/mi.pdf).

### 2020-04-15 3.Thesis/1.SimulationStudy/Writeup/manuscript.tex
Include the acf into simulation again to show differences

Missing data problems are ubiquitous and pervasive. \textbf{or}
Everybody who works with person data is likely to run into a missing
data problem.\\
Missingness is problematic because statistical inference cannot be
performed on the whole data without employing \emph{ad hoc} solutions
(e.g., list-wise deletion), which may yield wildly invalid results
\citep{buur18}. A popular method to obtain statistically valid results
despite of missing data is `multiple imputation' \citep[MI;][]{rubin76}.
The multiple imputation method can be roughly divided into 3 steps:
First, each missing value in an incomplete dataset is `imputed' (i.e.,
filled in) several times by some algorithmic procedure. Next, the
analysis of scientific interest is performed on each of the completed
datasets. And finally, the results of these analyses are pooled into a
single aggregated statistic.

>>>>

Missing data problems are ubiquitous and pervasive in the social and
behavioral sciences. If a dataset contains just one missing value for a
variable of interest, statistical inferences are undefined and will not
yield results. Incomplete observations are therefore ignored by default
in many statistical packages (i.e., list-wise deletion is employed).
Unfortunaltely, this \emph{ad hoc} solution may yield wildly invalid
results \citep{buur18}. An alternative is to `impute' (i.e., fill in)
every missing value in the incomplete dataset. With imputation
techniques, one or several completed datasets are created, on which
statistical inferences can be performed. The case with several completed
datasets is known as `multiple imputation' \citep[MI;][]{rubin76}, and
requires and additional step to pool the results accross imputations.
Because this pooled result will reflect the uncertainty in the data due
to missingness, it is in many cases an unbiased and confidence valid
estimate of the true (but missing) data inference \citep{buur18}. A
popular method to obtain imputations is to use the `Multiple Imputation
by Chained Equations' algorithm, shorthand `MICE'. MICE is an iterative
algorithmic procedure to draw imputations from the posterior predictive
distribution of the missing values. This introduces a potential threat
to the validity of the imputations: what if the algorithm has not
converged? Are the implications then to be trusted? And can we rely on
the inference obtained on the completed data? These are open questions,
because the convergence of iterative imputation algortihms has not been
systematically studied \citep{buur18}. In this paper, we investigate the
convergence properties of iterative imputation procedures by means of
model based simulation in \texttt{R} list(title = ``R: A Language and
Environment for Statistical Computing'', author = list(list(given = ``R
Core Team'', family = NULL, role = NULL, email = NULL, comment = NULL)),
organization = ``R Foundation for Statistical Computing'', address =
``Vienna, Austria'', year = ``2020'', url =
``\url{https://www.R-project.org/}'').

Rubin's rules \citeyearpar{rubin76} may be used to reflect the
uncertainty in the inferences due to missingness by pooling the results
across `imputations'. A popular method to obtain imputations
completions, the analysis of scientific interest can be performed. It
has been , or to complete these observations by imputing the missing
values. Imputation is the method of filling in

by which That is, unless ad hoc solutions are employed (e.g., list-wise
deletion), which can yield invalid results. contains missing values,
statistical analyses cannot be performed without only be performed a
dataset contains missing values, statistical analyses can statistical
inferences cannot be performed\\
Missingness is problematic because statistical inference cannot be
performed on the whole data without employing \emph{ad hoc} solutions
(e.g., list-wise deletion), which may yield wildly invalid results
\citep{buur18}. A popular method to obtain statistically valid results
despite of missing data is `multiple imputation' \citep[MI;][]{rubin76}.
The multiple imputation method can be roughly divided into 3 steps:
First, each missing value in an incomplete dataset is `imputed' (i.e.,
filled in) several times by some algorithmic procedure. Next, the
analysis of scientific interest is performed on each of the completed
datasets. And finally, the results of these analyses are pooled into a
single aggregated statistic.
